Notes on Control Plane Inference

### 2021-05-26 inference.tex
more lemmas

% commands
\newcommand{\assert}{\mathop{\mathsf{assert}}}
\newcommand{\assume}{\mathop{\mathsf{assume}}}
\newcommand{\apply}{\mathsf{apply}}
\newcommand{\choiceop}{\rotatebox[origin=c]{90}{$\sqsubset\!\!\!\sqsupset$}}
\newcommand{\choice}{\mathbin{\choiceop}}
\DeclareMathOperator*{\bigchoice}{\scalerel*{\choiceop}{\sum}}
\newcommand{\havoc}[1]{\mathop{\mathsf{havoc}}#1}
\newcommand{\SKIP}{\mathsf{skip}}

@@

\begin{definition}[Total State]
  \label{def:total-state}
  A state $\sigma$ is total if for every $t \in \dom(\sigma)$, the following
  formula is valid:
  \[\bigvee_{r \in \sigma(t.\id)} t.\keys = r.\matches \]
\end{definition}

Henceforth we assume that all states are total.

\todo[inline]{Can be Removed(?)}

@@

@@

\begin{lemma}[Disjoint Commutativity]
  \label{lem:disj-comm}
  For $c_1,c_2 \in \Cmd$, and $\sigma : \Value \to \Row^*$ if $\fvs(c_1) \cap
  \fvs(c_2) = \emptyset$, $ \forall \pkt \in \Pkt. \denote{c_1;c_2}^\sigma\pkt =
  \denote{c_2;c_1}^\sigma \pkt$.
\end{lemma}
\begin{proof}
  %% Let $c_1,c_2$, $\sigma$, and $\pkt$ be given as above. Assume $\fvs(c_1) \cap
  %% \fvs(c_2) = \emptyset$. Compute:
  %% \[
  %% \begin{array}{ll}
  %%   \denote{c_1;c_2}^\sigma \pkt =

  %% \end{array}
  %% \]
  \todo[inline]{prove}
\end{proof}

@@

\begin{lemma}{Subset Implies Projected Subset}
  \label{lem:subset-proj}
  For $P,Q \subseteq \Pkt$, and $ V \subseteq \Var$,
  If $P \subseteq Q$ then $P \subseteq_V Q$.
\end{lemma}

\begin{lemma}{Equality Implies Projected Equality}
  \label{lem:equal-proj}
  For $P,Q \subseteq \Pkt$, and $ V \subseteq \Var$,
  If $P = Q$ then $P =_V Q$.
\end{lemma}

@@

@@

\section{Data-Driven Precondition Inference?}

>>>>

% commands
\newcommand{\assert}{\mathop{\mathsf{ast}}}
\newcommand{\assume}{\mathop{\mathsf{asm}}}
\newcommand{\apply}{\mathsf{apply}}
\newcommand{\choiceop}{\rotatebox[origin=c]{90}{$\sqsubset\!\!\!\sqsupset$}}
\newcommand{\choice}{\mathbin{\choiceop}}
\DeclareMathOperator*{\bigchoice}{\scalerel*{\choiceop}{\sum}}
\newcommand{\havoc}[1]{\mathop{\mathsf{hvc}}#1}
\newcommand{\SKIP}{\mathsf{skip}}

@@

%% \begin{definition}[Total State]
%%   \label{def:total-state}
%%   A state $\sigma$ is total if for every $t \in \dom(\sigma)$, the following
%%   formula is valid:
%%   \[\bigvee_{r \in \sigma(t.\id)} t.\keys = r.\matches \]
%% \end{definition}

%% Henceforth we assume that all states are total.

%% \todo[inline]{Can be Removed(?)}

@@

\subsection{Definitions}

\begin{lemma}[Subset Implies Projected Subset]
  \label{lem:subset-proj}
  For $P,Q \subseteq \Pkt$, and $ V \subseteq \Var$,
  If $P \subseteq Q$ then $P \subseteq_V Q$.
\end{lemma}

\begin{lemma}[Equality Implies Projected Equality]
  \label{lem:equal-proj}
  For $P,Q \subseteq \Pkt$, and $ V \subseteq \Var$,
  If $P = Q$ then $P =_V Q$.
\end{lemma}

@@

\begin{lemma}[Pseudo-Parametricity]
  \label{lem:parametricity}
  For $c \in \Cmd$, $\sigma : \State$, $y \in \Var$, $\pkt \in \Pkt$, if $y
  \not\in \fvs(c)$, then $\forall \pkt' \in \denote{c}^\sigma \pkt$, $\pkt'(y) =
  \pkt(y)$.
\end{lemma}

\begin{proof}
  Let $\sigma : \State $, $y \in \Var$, and $\pkt \in \Pkt$ be given.
  Proceed by induction on $c$.
  \begin{enumerate}[align=left]
  \item[($c = x := e$)] Assume $y \not \in \fvs(x:=e)$, that is, $y \neq x$.
    Compute $\denote{x:=e}^\sigma \pkt = \{\pkt\{ x \mapsto \edenote e
    \pkt\}\}$. Observe that $\pkt\{ x \mapsto \edenote e \pkt\}(y) = \pkt(y)$
    since $y \neq x$. The result follows.

  \item[($c = \havoc x$)] Assume $y \not \in \fvs(\havoc x)$, that is $y \neq
    x$. Compute $\denote{\havoc x}^\sigma\pkt = \{\pkt\{x \mapsto v\} | v \in
    \Value\}$. let $v \in \Value$ be arbitrary. Observe that since $x \neq y$,
    $\pkt\{x \mapsto v\}(y) = \pkt(y)$. The result follows.

  \item[($c = \assume b$)] Assume $y \not \in \fvs(\assume b)$, that is $y
    \not\in \bfvs(b)$. Compute $\denote{\assume b}^\sigma \pkt = \{\pkt \mid
    \pkt \models b\}$. If $\pkt \not\models b$ the result holds vacuously.
    Otherwise the result follows from $\pkt(y) = \pkt(y)$.

  \item[$(c = \assert b)$] \textit{sim}.

  \item[$(c = c_1;c_2)$] Assume $y \not \in \fvs(c_1;c_2)$. Consequently $y
    \not\in \fvs(c_1)$, and $y \not \in \fvs(c_2)$. By IH$(c_1)$, $\forall
    \pkt_1 \in \denote{c_1}^\sigma(\pkt)$, $\pkt_1(y) = pkt(y)$. Pick an
    arbitrary $\pkt_1 \in \denote{c_1}^\sigma(\pkt)$. By IH$(c_2)$, $\forall
    \pkt_2 \in \denote{c_1}^\sigma(\pkt_1)$, $\pkt_2(y) = \pkt_1(y)$. By
    transitivity, we have $\pkt_2(y) = \pkt(y)$. Conclude that $\forall \pkt_2
    \in {\denote{c_2}^\sigma}^\dagger (\denote{c_1}^\sigma \pkt)$, $\pkt_2(y) =
    \pkt_1(y)$.

  \item[$(c = c_1 \choice c_2)$] Assume $y \not \in \fvs(c_1 \choice c_2)$.
    Consequently, $y \not \in \fvs(c_1)$ and $y \not \in \fvs(c_2)$.
    By IH$(c_i)$, $\forall \pkt' \in \denote(c_i)^\sigma\pkt$, for $i = 1,2$.
    By definition of $\denote{\cdot}$, the result follows.

  \item[$(t.\apply())$] Assume $y \not\in \fvs(t.\apply())$.
    There are two cases. Either $\pkt$ hits in the table or it misses.
    \begin{enumerate}[align=left]
    \item[\textit{Case 1}.] Assume that there is a row $r$ such that $\pkt
      \models t.\matchrow(r)$. Then $\denote{t.\apply()}^\sigma \pkt =
      \denote{t.\action[r.\action](r.\data)} \pkt$. The result follows by the
      IH.

    \item[\textit{Case 2}.] Conversely, there is no such row, in which case
      $\denote{t.\apply()}^\sigma\pkt = \{\pkt\}$, in which case the result is
      trivial.
    \end{enumerate}
  \end{enumerate}
\end{proof}

\begin{lemma}[Unused-Variable Expression Noninterference]
  \label{lem:exp-param-noninter}
  For $e \in \BVExpr$, $\pkt \in \Pkt$,
  $\{v_1,v_2\} \subseteq \Value$, $y \in \Var \setminus \efvs(e) $,
  $\edenote{e}\pkt\{y \mapsto v_1\} = \edenote{e}pkt\{y \mapsto v_2\}$.
\end{lemma}
\begin{proof}
  Let $\pkt$, $v_1$, $v_2$, and $y$ be given as above. Proceed by induction on
  $e$.
  \begin{enumerate}[align=left]
    \item[$(e = v)$] Compute: $\edenote{v}\pkt\{y
      \mapsto v_1\} = v = \edenote{v}pkt\{y \mapsto v_2\}$.
    \item[$(e = x)$] Since $y \not\in \efvs(x)$, conclude $y \neq x$. Compute:
      $\edenote{x}\pkt\{y \mapsto v_i\} = \pkt\{y \mapsto v_i\}(x) = \pkt(x) $
      for $i = 1,2$. The result follows.

    \item[$(e = e_1 \binop e_2)$]
      Note that $\efvs(e_i) \subseteq \efvs(e_1 \binop e_2)$ for $i = 1,2$. Compute
      \[\begin{array}{ll}
      \edenote{e_1 \binop e_2}\pkt\{y \mapsto v_1\} \\
      = \edenote{e_1}\pkt{y \mapsto v_1} \binop \edenote{e_2}\pkt\{y \mapsto v_1\} \\
      = \edenote{e_1}\pkt{y \mapsto v_2} \binop \edenote{e_2}\pkt\{y \mapsto v_2\} & \text{By IH}(e_i), i = 1,2\\
      = \denote{e_1 \binop e_2}\pkt\{y \mapsto v_2\}
      \end{array}
      \]

    \item[$(e = \unop e)$]
      Note that $\efvs(e) \subseteq \efvs(\unop e)$. Compute:
      \[\begin{array}{ll}
      \edenote{\unop e}\pkt\{y \mapsto v_1\} \\
      = \unop \edenote{e}\pkt\{y \mapsto v_1\} \\
      = \unop \denote{e}\pkt\{y \mapsto v_2\} & \textit{by IH}(e)\\
      = \denote{\unop e}\pkt\{y \mapsto v_2\}
      \end{array}\]
  \end{enumerate}
\end{proof}

\begin{lemma}[Unused-Variable Boolean Noninterference]
  \label{lem:bool-noninter}
  For $b \in \BExpr$, $\pkt \in \Pkt$,
  $\{v_1,v_2\} \subseteq \Value$, $y \in \Var \setminus \bfvs(b) $,
  $\pkt\{y \mapsto v_1\} \models b \iff pkt\{y \mapsto v_2\} \models b$.
\end{lemma}

\begin{proof}
  Let $\pkt$, $v_1$, $v_2$, and $y$ be given as above.
  Proceed by induction on $b$:
  \begin{enumerate}[align=left]
  \item[($b = \FALSE$)] Trivial.
  \item[($b = b_1 \Rightarrow b_2$)] Since $y \not \in\bfvs(b_1 \Rightarrow b_2) =
    \bfvs(b_1) \cup \bfvs(b_2)$, $y \not \in \bfvs(b_i)$, for $i = 1,2$. Compute
    as follows:
    \[\begin{array}{ll}
    \pkt\{y \mapsto v_1\} \models b_1 \Rightarrow b_2 \\
    \iff \pkt\{y \mapsto v_1\} \not\models b_1 \text{ or } \pkt\{y \mapsto v_1\} \models b_2 \\
    \iff \pkt\{y \mapsto v_2\} \not\models b_1 \text{ or } \pkt\{y \mapsto v_2\} \models b_2
    & \text{by IHs since } y \not \in \bfvs(b_i), i = 1,2 \\
    \iff \pkt\{y \mapsto v_2\} \models b_1 \Rightarrow b_2
    \end{array}\]

  \item[$(b = e_1 = e_2)$] Since $y \not \in\bfvs(e_1 = e_2) = \efvs(e_1) \cup
    \efvs(e_2)$, $y \not \in \bfvs(e_i)$ for $i =1,2$. Compute as follows:
    \[\begin{array}{ll}
    \pkt\{y \mapsto v_1\} \models e_1 = e_2
    \iff \edenote{e_1}\pkt\{y \mapsto v_1\} = \edenote{e_2}\pkt\{y \mapsto v_1\} \\
    \iff\edenote{e_1}\pkt\{y \mapsto v_2\} = \edenote{e_2}\pkt\{y \mapsto v_2\} & \text{by Lemma~\ref{lem:exp-param-noninter}} \\
    \iff \pkt\{y\mapsto v_2\} \models e_1 = e_2
    \end{array}
    \]
  \end{enumerate}
\end{proof}

\begin{lemma}[Unused Variable Noninterference]
  \label{lem:param-noninter}
  For $c \in \Cmd$, $\sigma : \State$, $\pkt \in \Pkt$,
  $\{v_1,v_2\} \subseteq \Value$, $y \in \Var \setminus \fvs(c)$,
  $\denote{c}^\sigma \pkt\{y \mapsto v\} =_{\Var \setminus \{y\}}\denote{c}^\sigma\pkt\{y \mapsto v\}$.
\end{lemma}

\begin{proof}
  Let $\sigma$, $v_1$, $v_2$, and $y$ be given.
  Proceed by induction on $c$, leaving $\pkt$ and $x$ general.
  \begin{enumerate}[align=left]
  \item[$(c = z := e)$]
    Let $\pkt$ be given. Since $y \not\in \fvs(z := e)$ we know that $y \neq z$.
    Compute:
    \[\begin{array}{ll}
    \pkt' \in \denote{z:=e}^\sigma (\pkt\{y \mapsto v\}) \\
    \iff \pkt' \in \{\pkt\{y \mapsto v\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v\})\}\}\\
    \iff \pkt' = \pkt\{y \mapsto v\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v\})\}
    \end{array}
    \]
    Similarly, compute:
    \[\begin{array}{ll}
    \pkt'' \in \denote{z:=e}^\sigma (\pkt\{y \mapsto v'\})\\
    \iff \pkt'' \in \{\pkt\{y \mapsto v'\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v'\})\}\}\\
    \iff \pkt'' = \pkt\{y \mapsto v'\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v'\})\}
    \end{array}
    \]

    We need to show that
    \[\forall x \in \Var \setminus \{y\}.
    \pkt\{y \mapsto v'\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v'\})(x)
    = \pkt\{y \mapsto v\}\{z \mapsto \edenote{e}(\pkt\{y\mapsto v\})\}(x)
    \]

    Equivalently we can show
    \[\forall x \in \Var \setminus \{y\}.
    \pkt\{z \mapsto \edenote{e}(\pkt\{y\mapsto v'\})(x)
    = \pkt\{z \mapsto \edenote{e}(\pkt\{y\mapsto v\})\}(x)
    \]

    Let $x \neq y$ be given. The result is immediate if $x \neq z$. Otherwise we
    must show that $\edenote{e}(\pkt\{y \mapsto v\}) = \edenote{e}(\pkt\{y
    \mapsto v'\})$. Since $y \not \in \efvs(e)$, this follows by
    Lemma~\ref{lem:exp-param-noninter}.

  \item[$(c = \havoc z)$] Let $\pkt$ by given. Since $y \not \in \fvs(\havoc z)$, we know $y \neq z$.
    Compute:
    \[\begin{array}{ll}
    \denote{\havoc z}^\sigma\pkt\{y \mapsto v_1\} \\
    = \{\pkt\{y \mapsto v_1\}\{z \mapsto v_z\} \mid v_z \in \Value\}\\
    =_{\Var \setminus \{y\}} \{\pkt\{y \mapsto v_2\}\{z \mapsto v_z\} \mid v_z \in \Value\} \\
    = \denote{\havoc z}^\sigma\pkt\{y \mapsto v_2\}
    \end{array}
    \]

  \item[$(c = \assume b)$]
    Let $\pkt$ by given. Since $\bfvs(b) \subseteq \fvs(\assume b)$, $y \not\in \bfvs(b)$.
    Compute \[
    \begin{array}{ll}
      \denote{\assume b}^\sigma \pkt\{y \mapsto v_1\} \\
      = \{\pkt \mid \pkt\{y \mapsto v_1\} \models b\}\\
      =_{\Var \setminus \{y\}} \{\pkt \mid \pkt\{y \mapsto v_2\} \models b\} & \text{by Lemma~\ref{lem:bool-noninter}, since }y \not\in \bfvs(b) \\
      = \denote{\assume b}^\sigma \pkt\{y \mapsto v_2\}
    \end{array}
    \]

    \item[$(c = \assert b)$] \textit{sim}.
    \item[$(c = c_1;c_2)$] Observe that $\fvs(c_i) \subseteq \fvs(c_1;c_2)$ for
      $i = 1,2$, which enables both IHS. Observe that:
      \[\begin{array}{ll}
      \denote{c_1;c_2}^\sigma\pkt\{y \mapsto v_1\}
      = {\denote{c_2}^\sigma}^\dagger\left(\denote{c_1}^\sigma\pkt\{y\mapsto v_1\}\right) \\
      \end{array}
      \]
      If $c_1$ denotes $\error$, the result is immediate, so assume otherwise.

      Let $\{i,j\} \subseteq \{1,2\}$. Let $\pkt_i \in
      \denote{c_1}^\sigma\pkt\{y \mapsto v_i\}$ be arbitary. By
      Lemma~\ref{lem:parametricity}, we know that $\pkt_i(y) = v_i$. Further
      IH$(c_1)$ tells us that there is some $\pkt_j \in
      \denote{c_1}^\sigma\pkt\{y\mapsto v_j\}$ such that $\pkt_1 =_{\Var
        \setminus \{y\}} \pkt_j$. Again, IH$(c_1)$ tells us that $\pkt_j(y) =
      v$. Consequently, we can write $\pkt_i$ and $\pkt_j$ in terms of a another
      packet $\pkt'$, i.e $\pkt_k = \pkt'\{y \mapsto v_k\}$ for $k = i,j$. Then
      by IH$(c_2)$, $\denote{c_2}^\sigma\pkt'\{y \mapsto v_i\} =_{\Var \setminus \{y\}}
      \denote{c_2}^\sigma \pkt'\{y \mapsto v_j\}$. Conclude that
      $\denote{c_1}^\sigma\pkt\{y \mapsto v_i\} \subseteq_{\Var \setminus \{y\} }
      \denote{c_1}^\sigma\pkt\{y \mapsto v_j\}$.

      We just proved that the projected subset is bidrectional, i.e. that
      $\denote{c_1}^\sigma\pkt\{y \mapsto v_1\} =_{\Var \setminus\{y\}}
      \denote{c_1}^\sigma\pkt\{y \mapsto v_2\}$

    \item[$(c = c_1 \choice c_2)$] Observe that $\fvs(c_i) \subseteq \fvs(c_1 \choice c_2)$ for
      $i = 1,2$, which enables both IHs. Compute as follows
      \[\begin{array}{ll}
      \denote{c_1 \choice c_2}^\sigma\pkt\{y \mapsto v_1\} \\
      = \denote{c_1}^\sigma\pkt\{y \mapsto v_1\} \cup \denote{c_2}^\sigma\pkt\{y \mapsto v_1\} \\
      =_{\Var \setminus \{y\}} \denote{c_1}^\sigma\pkt\{y \mapsto v_2\} \cup \denote{c_2}^\sigma\pkt\{y \mapsto v_2\}
      & \text{by IHs} \\
      = \denote{c_1 \choice c_2}
      \end{array}
      \]

    \item[$(c = t.\apply())$]
      There are two cases, corresponding to hitting and missing.
      \begin{enumerate}[align=left]
      \item[\textit{Case 1}.] Assume there is some $r \in \sigma(t)$ such that
        $\pkt\{y \mapsto v_1\} \models t.\matchrow(r)$, and that $r$ is the
        minimally-indexed such row. Then by Lemma~\ref{lem:bool-noninter},
        $\pkt\{y \mapsto v_2\} \models t.\matchrow(r)$, and is the minimally indexed such row. Then, since
        $\fvs(t.\action[r.\action](r.\data)) \subseteq \fvs(t.\apply())$, the
        result follows by IH$(t.\action[r.\action](r.\data))$.

        \item[\textit{Case 2}.] Assume there is no $r \in \sigma(t)$ such that
          $\pkt\{y \mapsto v_1\} \models t.\matchrow(r)$. Then by
          Lemma~\ref{lem:bool-noninter}, there is now $r \in \sigma(t)$ such
          that $\pkt\{y \mapsto v_2\} \models t.\matchrow(r)$. Then by definitions:
          \[\denote{t.\apply()}^\sigma \pkt\{y \mapsto v_1\} = \{\pkt\{y \mapsto v_1\}\} = \{\pkt\{y \mapsto v_2\}\} = \denote{t.\apply()}^\sigma \pkt\{y \mapsto v_2\}\]
          The result follows by Lemma~\ref{lem:equal-proj}.
      \end{enumerate}

\begin{lemma}[Disjoint Commutativity]
  \label{lem:disj-comm}
  For $c_1,c_2 \in \Cmd$, and $\sigma \in \State$ if $\fvs(c_1) \cap
  \fvs(c_2) = \emptyset$, $ \forall \pkt \in \Pkt. \denote{c_1;c_2}^\sigma\pkt =
  \denote{c_2;c_1}^\sigma \pkt$.
\end{lemma}
\begin{proof}
  \todo[inline]{todo}
\end{proof}

@@

@@

%% \begin{lemma}[Universal Quantifier Weakness]
%%   For $b \in \BExpr$ and $V \subseteq \bfvs(b)$, $\forall V. b$ is the weakest
%%   formula on variables $\bfvs(b) \setminus V$ that implies $b$.
%% \end{lemma}
%% \begin{proof}
%%   Elementary.
%% \end{proof}

@@

  \todo[inline]{weakest?}
\end{proof}

\clearpage
\section{Data-Driven Precondition Inference?}
